---
title: "Matrix algebra"
author: "Chrissy Rochus"
date: "`r Sys.Date()`"
output: pdf_document
---

This Rmarkdown file corresponds to the examples in the lecture on "Matrix algebra". 

# Special matrices

Two of the special square matrices can be made with the base R function "diag()".  The first is the identity matrix (off diagonal are zeros, diagonals are all 1s).  Since it is square, within the brackets, include the number of columns/rows needed.

```{r}
# For an identity matrix with 3 rows and 3 columns
diag(3)
```

Diagonal matrices can also be made with the "diag()" function.  Within the brackets, provide the values for the diagonal of the square matrix.

```{r}
# For a diagonal matrix with a diagonal of 5, -3, and 1
diag(c(9, 3, -1))
```

# Addition

Two matrices can be added together if they have the same number of rows and columns.

```{r}

A <- matrix(c(9,-11,2,5),nrow=2,ncol=2)
A

B <- matrix(c(13,8,-7,4),nrow=2,ncol=2)
B

A + B

```

# Subtraction

A matrix can be subtracted from another if they have the same number of rows and columns.

```{r}
A - B
```

# Multiplication

Two matrices can be multiplied if the number of columns in the first matrix equals the number of rows in the second.

```{r}
E <- matrix(c(-1,6,4,5,-2,-3,3,4,6),nrow=3,ncol=3)
E

G <- matrix(c(4,2,-6,-5,3,1),nrow=3,ncol=2)
G

# For multiplying scalars in R, use the operator '*', for matrices, use '%*%'
E%*%G

```
# Dot product

The sum of the products of two equal length vectors.

```{r}

a <- c(7,-2,1)
a

b <- c(3,5,-4)
b

# When vectors are stored as values, use '%*%' to multiply them to get the dot product

a%*%b

```
# Multiplying a scalar and a matrix

Multiplying a matrix with a scalar, multiplies each element of the matrix with the scalar.

```{r}

4*E

# In which order the scalar and matrix are multiplied doesn't matter
# 4*E is equal to E*4

E*4

```

# Direct (Kronecker) product of matrices

For a matrix, E (order r x s), and H (order m x n), the direct product, has an order of rm x sn.

```{r}

H <- matrix(c(4,2,-5,3),nrow=2,ncol=2)

kronecker(E, H)

# The direct product of E and H is not equal to the direct product of
# H and E
kronecker(H, E)

```

# Hardamard product

For two matrices of the same dimension, the Hardamard product has the same dimensions.

```{r}

X <- matrix(c(7,2,-4,2,3,0,-4,0,-9),nrow=3,ncol=3)
X

E

E*X

# The Hardamard product of E and X is equal to the Hardamard product
# of X and E
X*E

```

# Transpose

For a matrix J, with order i x k, the transpose, J’ has order i x k.

A matrix is not equal to its transpose unless it is a symmetric matrix (like matrix W).


```{r}

J <- matrix(c(5,-1,6,3,4,-2),nrow=3,ncol=2)

J

t(J)

W <- matrix(c(7,2,-4,2,3,0,-4,0,-9),nrow=3,ncol=3)

W

t(W)

```

# Dot product, part 2

When vectors are stored as matrices (with row = 1), use '%*%' to multiply one with the transpose of the other to get the dot product

```{r}

a <- matrix(c(7,-2,1),nrow=1,ncol=3)
a

b <- matrix(c(3,5,-4),nrow=1,ncol=3)
b

a%*%t(b)

```

# Trace

The sum of the diagonal elements of a matrix.

```{r}

sum(diag(W))

```

# Inversion

When an inverse of a matrix is multiplied by the matrix (W in this example), an identity matrix (I) is the product.

Only square matrices can be inverted.

For a diagonal matrix, Y, the inverse is the reciprocal of the diagonal elements.

```{r}

solve(W)%*%W

Y <- diag(c(5,-10,2))
Y

solve(Y)

# Z is a 2x2 matrix example

Z <- matrix(c(3,2,-1,5),nrow=2,ncol=2)

solve(Z)
  
```

# Rank

The number of linearly independent rows or columns.

A square matrix with rank equal to the number of rows or columns is full rank.

```{r}
N <- matrix(c(2,5,-3,5,3,2,2,1,1),nrow=3,ncol=3)

N

SVD <- svd(N) # singular value decomposition of the matrix

sum(SVD$d > 1e-10) # rank equals the number of nonzero singular values

```

# Generalised inverse

There are many different ways to calculate a generalised inverse, so this inverse may not be the same as the one calculated using pen and paper for this example.

```{r}

# If the "MASS" package hasn't already been installed,
# use the following line to install it (remove # and run)
#install.packages("MASS")
library(MASS)

ginv(N)

# There are different ways to calculate the generalised inverse, but this is 
# one way to test if the generalised inverse calculated is correct: 
# the matrix (N) multiplied by its inverse, multiplied by N is equal to N.
N %*% ginv(N) %*% N

N

```

# Eigenvalues and eigenvectors

Eigenvalues are useful in simplifying multivariate evaluations when transforming data.

```{r}

res = eigen(N)

# Eigen values
res$values

# Eigen vectors
res$vectors

```

# Cholesky decomposition

The Cholesky decomposition of a matrix (K) is a lower triangular matrix (V).  We can derive V from K.


```{r}

K <- matrix(c(9,3,-6,3,5,0,-6,0,21),nrow=3,ncol=3)
K

# Use chol() to find the transpose of V
V <- t(chol(K))
V

# K is equal to V%*%t(V)
V%*%t(V)

```
